
title: "Roots traits"
author: "Ferreira-Arruda"
date: "2024-01-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Load necessary libraries
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(FactoMineR)
library(factoextra)
library(glue)
library(vegan)
library(ggpubr)
library(FactoMineR)
library(ggrepel)
library(ggExtra)
library(stringr)

# Read the data
data <- read.csv("./data/pairs_traits.csv", sep = ";", dec = ",")
head(data)
nitro.d<- read.csv("./data/nitro.csv",sep = ";", dec = ",")

data$sample_id <- paste(data$tag_id, data$root_type, sep = "_")
#\\d{4}: Matches exactly 4 consecutive digits.
#[a-z]{2}: Matches an underscore followed by exactly two lowercase

nitro.d$sample_id<- str_extract(nitro.d$Originalbez, "(?<=SRI_).*?(?=_[^_]*$|$)")
nitro.d<- select(nitro.d,Originalbez, sample_id, Ntotal.mmol.g.)
nitro.d<- na.omit(nitro.d)
nitro.d$sample_id[which(nitro.d$Originalbez == "54_SRI_MT_ab_O")]<- "MT_54_ab"
nitro.d$sample_id[which(nitro.d$Originalbez == "55_SRI_MT_ab_")]<- "MT_55_ab"
nitro.d$sample_id[which(nitro.d$Originalbez == "54_SRI_MT_tr_O")]<- "MT_54_tr"
nitro.d$sample_id[which(nitro.d$Originalbez == "55_SRI_MT_tr_")]<- "MT_55_tr"



nitro.d$sample_id <- sub("^0+", "", nitro.d$sample_id)


data<- left_join(data, nitro.d, by = "sample_id")
data<- data %>% select(., -Originalbez)
data<- data[!is.na(data$Ntotal.mmol.g.), ]

#merge OO with CRTL because what am I gonna do with 2 sample? 

data$system[which(data$system == "OO")]<- "CTRL"


```

# Merging TRA and ABS roots

```{r}

roots_mer <- data %>%
  group_by(tag_id) %>% select(-plot) %>%
  summarize(across(where(is.numeric), ~ sum(. * dry_weight_g, na.rm = TRUE) / sum(dry_weight_g, na.rm = TRUE), .names = "weighted_avg_{col}"))


roots_mer<- unique(left_join(roots_mer, select(data, tag_id, species, system), by = "tag_id"))
roots_mer <- roots_mer %>%
  filter(!if_any(everything(), is.nan))
```

## Keeping only the traits we are interested

```{r}
roots.m.sub<- roots_mer %>% select(tag_id, species,system, weighted_avg_ratio_abs_tra, weighted_avg_tips_lenghts, 
                                   weighted_avg_AvgDiamWeight.mm., weighted_avg_RTD, weighted_avg_SRL, weighted_avg_SRA, weighted_avg_ratio_abs_tra, weighted_avg_Ntotal.mmol.g.)

colnames(roots.m.sub)<- c("tag_id", "species", "system", "Abs_Tra_ratio", "Tip_lenght", "Diameter",
                          "RTD", "SRL", "SRA", "Nitrogen")

```

# Oil palm trait space

## Only oil palms

```{r}
oil_alone<- roots.m.sub %>% filter(system == "OO" | system== "CTRL")
# Perform PCA
pca_result <- PCA(oil_alone[, -c(1:3)], scale.unit = TRUE, graph = FALSE)

# Plotting the PCA with different colors for 'OO' and 'CTRL' systems
fviz_pca_biplot(pca_result,col.ind = oil_alone$system,
                #palette = c("blue", "red"),
                #addEllipses = TRUE, ellipse.type = "confidence",
                repel = TRUE, # Avoid text overlapping
                arrow.size = 0.5, arrow.type = "closed",
                label = "none")
```

## With native trees

```{r}

# Perform PCA
oil_traits<- filter(roots.m.sub, species == "O")



pca_result_all <- PCA(oil_traits[, -c(1:3)], scale.unit = TRUE, graph = FALSE)

# Plotting the PCA with different colors for 'OO' and 'CTRL' systems
fviz_pca_biplot(pca_result_all,col.ind = oil_traits$system,
                #addEllipses = TRUE, ellipse.type = "confidence",
                repel = TRUE, # Avoid text overlapping
                arrow.size = 0.5, arrow.type = "closed",
                label = "none")

```

# Traits distributions

```{r}

library(ggridges)
# Filter for species classified as 'O'
species_o_data <- roots.m.sub %>% filter(species == "O")

# Convert necessary columns to numeric
# Assuming we are interested in the same columns as before

numeric_data <- species_o_data %>% 
  select(-tag_id, -species) %>% 
  mutate(across(-system, ~ as.numeric(gsub(",", ".", .))))

# Creating ridge plots for each trait across systems
# Replace `trait_name` with the name of each trait column you want to plot
for (trait_name in colnames(numeric_data)[2:length(numeric_data)]) {
  plot <- ggplot(numeric_data, aes_string(x = trait_name, y = "system", fill = "system")) +
        scale_fill_manual(values = c("#9DD6E0", "#C670E0", "#E0AA7B", "#B0D6B1", "#E09DC6")) +
    geom_density_ridges(alpha =0.8) +
    geom_boxplot(aes_string(y = 0.2, x = trait_name, group = "system"), 
                 alpha = 0.5, width = 1.5, outlier.shape = NA) +
    #scale_y_continuous(expand = expansion(add = c(0.1, 0.2))) +
    labs(title = paste(trait_name, "per System"),
         x = trait_name,
         y = "System") +
    theme_minimal() +
    theme(legend.position = "none")
  
  print(plot)
  # Save the plot
  ggsave(filename = paste0("figures/", trait_name, "_system_plot.png"),
         plot = plot, 
         width = 8, height = 6, dpi = 300)
}


for (trait_name in colnames(numeric_data)[2:length(numeric_data)]) {
  plot <- ggplot(numeric_data, aes_string(y = trait_name, x = "system", fill = "system")) +
    scale_fill_manual(values = c("#9DD6E0", "#C670E0", "#E0AA7B", "#B0D6B1", "#E09DC6")) +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(alpha = 0.2) +
    labs(title = paste(trait_name, "per System"),
         x = "System",
         y = trait_name) +
    theme_minimal() +
    theme(legend.position = "none")
  
  print(plot)
}



```

High SRA points out to a more acquisitive strategy. Higher proportion of fine roots. Could indicate a shift towars are more competitive strategy, increasing fast resource acquisition.

Lower Root tissue density also shows a shift to a more fast growin acquisitive strategy.

Over all for all the traits we analyzed, Palm oil seems to widen the range of trait values when its competing with the native species.

```{r}
# Perform PCA on this subset
pca_result <- PCA(numeric_data[, -1], scale.unit = TRUE, ncp = 5, graph = FALSE)

# Extract PCA scores (individuals' coordinates)
pca_scores <- as.data.frame(pca_result$ind$coord)

# Add the 'system' column back to the pca_scores dataframe for coloring
pca_scores$system <- numeric_data$system

# Creating ridge plots for the first few PCA dimensions
# Creating ridge plots for the first few PCA dimensions
for (pc in 1:3) { 
  pc_name <- paste0("Dim.", pc) # Creates 'Dim1', 'Dim2', etc.
  
  # Dynamically creating plot titles and axis labels
  plot_title <- paste("Distribution of PCA Dimension", pc, "Scores Across Systems")
  x_label <- paste("PCA Dimension", pc, "Score")
  
  # Using aes() instead of aes_string() for dynamic variable names in ggplot2
  plot <- ggplot(pca_scores, aes(x = get(pc_name), y = system, fill = system)) +
    geom_density_ridges(alpha =0.5) +
    labs(title = plot_title, x = x_label, y = "System") +
    theme_minimal() +
    theme(legend.position = "none")
  
  print(plot)
}

```

```{r}

PCAloadings <- as.data.frame(pca_result$var$coord) # extract PCA loadings
PCAloadings$trait <- rownames(PCAloadings)



my_theme1 <- theme( axis.title.x=element_text(colour="black",face="bold",size=15),
                    axis.title.y=element_text(colour="black",face="bold",size=15),
                    axis.text.x=element_text(colour=c("black"),face="bold",size=15),
                    axis.text.y=element_text(colour=c("black"),face="bold",size=15),
                    legend.position = c(0.15,0.15), legend.direction="vertical", 
                    legend.key = element_rect(fill="transparent"),
                    legend.key.size = unit(0.5,"line"),
                    legend.title = element_blank(),
                    legend.text = element_text(size=12, color="black"),
                    panel.background = element_rect(fill="transparent",colour="black"),
                    panel.grid.minor = element_blank(),
                    panel.border = element_rect(fill=NA,colour="grey"),
                    plot.title = element_text(size = 20))



pca_plot<- ggplot(pca_scores, aes(Dim.1, Dim.2, col = system)) + 
  scale_color_manual(values = c("#9DD6E0", "#C670E0", "#E0AA7B", "#B0D6B1", "#E09DC6", "#A09DC6")) +
  #stat_density_2d(geom = "polygon", contour = TRUE, aes(fill = after_stat(level)), 
                  #colour = "gray80", bins = 10, show.legend = F) + #bins is for the number of contours
  geom_jitter(alpha = 0.9, size =3.5) + # Display the points
  geom_text_repel(data = PCAloadings, aes(x = Dim.1*3.5, y = Dim.2*3.3, label = trait), size = 4, inherit.aes = F, 
                  segment.color = 'transparent', color = "black") + # scale the loadings
  geom_segment(data = PCAloadings, size = 0.8, inherit.aes = F, # Plots the loadings, i.e., traits
               aes(x = 0, xend = Dim.1*3, y = 0, yend = Dim.2*3),
               arrow = arrow(length = unit(0.5, "cm")),colour = "black") +
  #xlab(paste("PC1 ", round(qt$eig[1, 2]), "%")) + ylab(paste("PC2 ", round(qt$eig[2, 2]), "%")) +
  #scale_x_continuous(limits=c(-5,6)) +
  #scale_y_continuous(limits=c(-5,5)) +
  #my_theme1 + guides(colour= guide_legend(override.aes = list(size=4))) +
  theme_minimal() +
  ggtitle("Oil Palm Roots Trait Space")+
  theme(legend.position = "right")
pca_plot

ggsave(filename = paste0("figures/", "oil_palm_pc", ".png"),
         plot = pca_plot, 
         width = 8, height = 6, dpi = 300)

pca_plot_dens<-  ggMarginal(pca_plot, type = "density", margins = "both", size = 5, groupColour = TRUE, groupFill = TRUE)
pca_plot_dens

```

```{r}
library(grid)
library(gridExtra)


var_contributions <- as.data.frame(pca_result$var$contrib[, 1:3])

# Round the contributions to 3 decimal places
var_contributions <- round(var_contributions, 3)

# Rename the columns for clarity
colnames(var_contributions) <- c("Dim 1 (%)", "Dim 2 (%)", "Dim 3 (%)")

# Add the variable names as a new column
var_contributions$Variable <- rownames(pca_result$var$contrib)

# Sort by 'Dim 1 (%)' in descending order to show higher contributions first
var_contributions <- var_contributions[order(-var_contributions$`Dim 1 (%)`),]

# Create the table plot
table_plot <- tableGrob(var_contributions, rows = NULL, theme = ttheme_minimal())

# Plotting the table. Note: This is not a ggplot object but a gridExtra object.
grid.draw(table_plot)

```

```{r}

pca_all_sp <- PCA(roots.m.sub[, -c(1:3)], scale.unit = TRUE, ncp = 5, graph = FALSE)


# Extract PCA scores (individuals' coordinates)
pca_scores2 <- as.data.frame(pca_all_sp$ind$coord)
pca_scores2$system <- roots.m.sub$system
pca_scores2$species <- roots.m.sub$species



PCAloadings2 <- as.data.frame(pca_all_sp$var$coord) # extract PCA loadings
PCAloadings2$trait2 <- rownames(PCAloadings2)



my_theme1 <- theme( axis.title.x=element_text(colour="black",face="bold",size=15),
                    axis.title.y=element_text(colour="black",face="bold",size=15),
                    axis.text.x=element_text(colour=c("black"),face="bold",size=15),
                    axis.text.y=element_text(colour=c("black"),face="bold",size=15),
                    legend.position = c(0.15,0.15), legend.direction="vertical", 
                    legend.key = element_rect(fill="transparent"),
                    legend.key.size = unit(0.5,"line"),
                    legend.title = element_blank(),
                    legend.text = element_text(size=12, color="black"),
                    panel.background = element_rect(fill="transparent",colour="black"),
                    panel.grid.minor = element_blank(),
                    panel.border = element_rect(fill=NA,colour="grey"),
                    plot.title = element_text(size = 20))



pca_plot2 <- ggplot(pca_scores2[-48,], aes(Dim.1, Dim.2, col = system)) + 
  scale_color_manual(values = c("#9DD6E0", "#C670E0", "#E0AA7B", "#B0D6B1", "#E09DC6", "#A09DC6", "#F0D4B9")) +
  geom_jitter(aes(shape = species), 
              alpha = 0.9, size = 3.5) + # Map shape directly to species
  scale_shape_manual(values = c("SHO" = 16, "PAR" = 16, "ARC" = 16, "PER" = 16, "O" = 15), guide = "none") + # Define shapes
  geom_text_repel(data = PCAloadings2, aes(x = Dim.1*3.5, y = Dim.2*3.5, label = trait2), 
                  size = 4, inherit.aes = F, 
                  segment.color = 'transparent', color = "black") + # Scale the loadings
  geom_segment(data = PCAloadings2, size = 0.8, inherit.aes = F, # Plot the loadings
               aes(x = 0, xend = Dim.1*3, y = 0, yend = Dim.2*3),
               arrow = arrow(length = unit(0.5, "cm")), colour = "black") +
  theme_minimal()+
  ggtitle("Oil Palm + Native trees Roots Trait Space")+
  theme(legend.position = "right")
pca_plot2
ggsave(filename = paste0("figures/", "all_trees_pc", ".png"),
         plot = pca_plot2, 
         width = 8, height = 6, dpi = 300)

pca_plot_dens2<-  ggMarginal(pca_plot2, type = "density", margins = "both", size = 5, groupColour = TRUE, groupFill = TRUE)
pca_plot_dens2

```
